\chapter{Inference rules}
\label{chapter:inference}
Inference rules define the valid proof steps in a proof system. The application supports inference rules of the form
\[
    \Inf[\text{Rule name}]{\text{Premises}}{\text{Conclusion}}
\]
which lets us derive the conclusion if all the premises hold. Inference rules can use placeholders to match any user-inputted term. Inference rules can also destructure user-inputted terms and reject user-inputted terms that cannot be destructured according to the inference rule. The premises and conclusion of an inference rule are specialised statements that carry structural information and should be parsed more like user-inputted terms than syntax rules. These specialised statements are called \textit{inference rule statements}, consisting of \textit{inference rule terms}, which are either placeholders or destructurings of user-inputted terms.

Like syntax rules, inference rules may be added, deleted, or replaced by the user. The web application must provide an interface for modifying inference rules, check whether the inference rules are defined sensibly, and be able to verify whether the user has applied an inference rule correctly.

\Cref{inference:syntax} defines the syntax in which a user expresses an inference rule in the application, describes the concept of \textit{inference rule terms}, and contrasts it with \textit{user-inputted terms} introduced in the previous chapter. \Cref{inference:ui} gives an overview of the parts of the user interface for modifying inference rules. \Cref{inference:parsing} describes the procedure for transforming an arbitrary set of inference rules into intermediate internal representations, in preparation for verifying applications of the inference rules in the derivation tree.

\section{Syntax of inference rules}
\label{inference:syntax}
An inference rule consists of:
\begin{itemize}
    \item A \textbf{name},
    \item One \textbf{conclusion} following the structure of a statement, and
    \item Zero or more \textbf{premises}, each following the structure of a statement
\end{itemize}
Inference rules are \textit{local}: checking whether an inference rule is applied correctly only requires information immediately above and below the step of applying the inference rule, but not any information further up or down the derivation tree.

\subsection{Terminology: inference rule terms and statements}
In this report, an \textit{inference rule statement} refers to one of the premises or the conclusion of an inference rule. A term appearing in an inference rule statement may not always be a valid term when given by the user. Consider the following syntax rule:
\[
    A, B \Coloneqq \varphi \alt (A \to B)
\]
In a derivation, the user can only input terms like $\varphi$, $(\varphi \to \varphi)$, and $((\varphi \to \varphi) \to \varphi)$. In an inference rule, however, an instance of a syntax rule can be represented by any of its placeholders and partially expanded versions of its definition. In this case, $A$, $B$, $(A \to B)$, $((A \to B) \to A)$, and even $((B \to \varphi) \to B)$ are all valid \textit{inference rule terms}. In this report, a \textit{(user-inputted) term} refers to an instance of a syntax rule that appears in derivations and an \textit{inference rule term} refers to an instance of a syntax rule that appears in an inference rule statement. All valid user-inputted are valid inference rule terms, and so the set of inference rule terms for any syntax definition with more than one non-statement rule is a strict superset of the set of user-inputted \textit{terms}\footnote{A syntax definition with only rules for statements does not have any placeholders, so the rule definitions cannot refer to any placeholders and must only consist of terminals and multisets. For such a syntax definition, the set of inference rule terms is equal to the set of terms.}.

\subsubsection{Placeholders in syntax rule definitions vs. inference rule statements}
Recall from \Cref{syntax:syntax} that in syntax rule definitions, placeholders for a rule can be used interchangeably. The following syntax rule is semantically identical to the example above:
\[
    A, B \Coloneqq \varphi \alt (A \to A)
\]
Even though $A$ appears twice in different positions of the same alternative, the two user-inputted terms that take their positions can be different. For example, $(\varphi \to (\varphi \to \varphi))$ is a valid instance of both versions of the syntax rule. However, the inference rule term $(A \to A)$ is compatible with a user-inputted term if and only if the two user-inputted sub-terms that take the positions of the two occurrences of $A$ are identical. In this case, $(\varphi \to (\varphi \to \varphi))$ is incompatible with the inference rule term $(A \to A)$ because the sub-terms $\varphi$ and $(\varphi \to \varphi)$ are different. The process of matching inference rule terms with user-inputted terms is explained in detail in \Cref{chapter:checking}.

\subsubsection{Multisets in inference rules}
\label{inference:multisets}
Consider the following syntax rules:
\begin{align*}
    \Gamma, \Delta &\Coloneqq \{ A \} \\
    A, B &\Coloneqq \varphi \alt (A \to B)
\end{align*}
On its own, $\Gamma$ represents a complete multiset. Now, consider the term $\Gamma, A$. $\Gamma$ represents a multiset with one element removed. The removed element is given the name $A$. Since an element can be removed, the user-inputted multiset must contain at least one element: trying to match the user-inputted term $\varnothing$ against the inference rule term $\Gamma, A$ results in an error. Note that $\Gamma$ can be empty. Analogously, it is possible to:
\begin{itemize}
    \item Match multiple multiset elements, such as $\Gamma, A, B$ and $\Gamma, A, A, A$.
    \item Match multiple multisets, such as $\Gamma, \Delta$.
    \item Match multiple multisets and multiset elements, such as $\Gamma, \Delta, A, B$.
\end{itemize}

It is also possible to match multiset elements with specific structures. The following are all valid inference rule terms:
\[
    \Gamma, (A \to B) \qquad \Gamma, (A \to (A \to A)) \qquad \Gamma, ((\varphi \to B) \to A)
\]
since $(A \to B)$, $(A \to (A \to A))$, and $((\varphi \to B) \to A)$ are all valid expansions of the definition of the second syntax rule, of which $A$ is a placeholder.

The order of the multiset placeholders and multiset element placeholders does not matter. This means a multiset term is identical to any of its permutations. The following terms are all identical:
\[
    \Gamma, \Delta, A, B \qquad \Gamma, B, \Delta, A \qquad A, B, \Delta, \Gamma \qquad \Delta, B, A, \Gamma \qquad \ldots
\]

Note that a placeholder (in this case, $\Gamma$ or $\Delta$) can represent a multiset if and only if at least one of its alternatives only consists of a multiset and no other tokens. For example, if the first syntax rule in the example above is changed to
\[
    \Gamma, \Delta \Coloneqq \{ A \}: B
\]
instead, things like $\Gamma, A$ and $\Gamma, \Delta, A$ are not valid inference rule terms for the modified syntax rule. 
\section{User interface}
\label{inference:ui}

\section{Parsing inference rules}
\label{inference:parsing}
The aim of this step is to generate parsers for parsing an inference rule statement into a list of \lstinline{Matchable} tokens.

\subsection{\texorpdfstring{\lstinline{Matchable}}{Matchable} tokens}
\lstinline{Matchable} tokens are to inference rules like ASTs are to terms and \lstinline{Token}s are to syntax rules. \lstinline{Matchable} tokens represent the structure of an inference rule statement. They are defined as follows:
\begin{lstlisting}
    class MatchableTerminal { constructor(readonly value: string) {} }

    class Name { constructor(readonly index: number, readonly name: string) {} }

    class MatchableNonTerminal {
        constructor(readonly index: number, readonly children: Matchable[]) {} }

    class MultisetElement { constructor(readonly tokens: Matchable[]) {} }

    class MatchableMultiset {
        constructor(
            readonly index: number,
            readonly elements: (Name | MultisetElement)[]
        ) {}
    }

    type Matchable = MatchableTerminal | Name | MatchableNonTerminal | MatchableMultiset;
\end{lstlisting}
The semantics of the \lstinline{Matchable} tokens are as follows:
\begin{itemize}
    \item \lstinline{MatchableTerminal}: match the string stored in the \lstinline{value} field.
    \item \lstinline{Name}: assign exactly one AST to \lstinline{name} and ensure this assignment is compatible with all other occurrences of the same \lstinline{name} in other statements of the same inference rule. The AST must represent an instance of the syntax rule at \lstinline{index}\footnote{The \lstinline{index} field can be deduced from the \lstinline{name} field since the placeholders are unique across all syntax rules. The \lstinline{index} field exists merely to simplify bookkeeping.}.
    \item \lstinline{MatchableNonTerminal}: match an inference rule term corresponding to the syntax rule at \lstinline{index}. The structure of the inference rule term is given by the \lstinline{children} field.
    \item \lstinline{MatchableMultiset}: match a multiset of elements given by the \lstinline{elements} field. A member of \lstinline{elements} can either be a \lstinline{Name} or a \lstinline{MultisetElement}. A \lstinline{Name} refers to a multiset (e.g. $\Gamma$ and $\Delta$ in the example in the previous section) while a \lstinline{MultisetElement} refers to a multiset element that must be matched from the user-inputted multiset term (e.g. $A$ and $(A \to B)$ in the previous section). The \lstinline{elements} field is represented as an (ordered) array instead of an unordered set purely for simpler bookkeeping in the matching algorithm.
\end{itemize}
\subsection{Procedure}
Parsing inference rule statements is extremely similar to parsing terms, as an inference rule statement can be thought of as an instance of a statement in a different context. As with generating term parsers, a delayed parser is created  for every syntax rule using the \lstinline{later} combinator. Each of these delayed parsers is instantiated by iterating over the tokens in every alternative of the syntax rule and examining the type of the token:
\begin{itemize}
    \item \lstinline{Terminal}: parse the string verbatim and wrap the result in a \lstinline{MatchableTerminal} token.
    \item \lstinline{NonTerminal}: either parse a placeholder of the rule corresponding to the rule number (i.e. the \lstinline{index} field of the \lstinline{NonTerminal}), or an expanded version of the definition of the syntax rule using the (delayed) parser corresponding to the rule. The former produces a \lstinline{Name} token while the latter produces a \lstinline{MatchableNonTerminal} token.
    \item \lstinline{Multiset}: either parse \lstinline{\varnothing} (corresponding to the empty set) or a comma-separated list of multiset elements, as described in \Cref{inference:multisets}. If placeholders for multisets are applicable here (i.e. the \lstinline{Multiset} token is the only token of an alternative of the syntax rule definition, and the syntax rule does not define a statement), the comma-separated list can contain both individual multiset elements and placeholders for multisets. Since the \lstinline{Multiset} token does not store the rule number of the rule it is part of, the rule number is passed as an argument to the function which generates inference rule parsers.
\end{itemize}